Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model

### 2023-05-31 analysis/paper/discussion.Rmd
fix: broken citekeys

<!-- Big summary/important statement for the science -->
Here, we have provided a method for exploring the incorporation of dietary 
starches into the mineral matrix of a dental calculus biofilm model. Our results show
that a very low proportion of the starches exposed to the biofilm during growth are
retained in the mineral matrix, and that the size of the starch granules
may affect the likelihood of incorporation. The proportions of starch granules 
(of all sizes) present in the extracted samples were similar across all treatments
(`r paste(min(perc_tbl$total), max(perc_tbl$total), sep = " to ")`), 
despite large differences in absolute granule counts between wheat 
(mean = `r format(sol_comb_out$total[5], big.mark = ",")`) 
and potato
(mean = `r format(sol_comb_out$total[4], big.mark = ",")`)
solutions.  
The absolute counts, however, differed more visibly between treatments and was
proportional with the total count of granules in the treatment solutions. Wheat 
and mixed solutions had the highest absolute mean count of starch granules, and 
also had the highest absolute mean count of starch granules extracted from the 
dental calculus
(Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).
This suggests that the starches that are more frequently consumed will be present
in higher quantities in the dental calculus, at least prior to inhumation and 
degradation in the burial environment. 
Despite the low proportion of granules recovered from the model calculus 
(`r paste(min(perc_tbl$total), max(perc_tbl$total), sep = " to ")`),
the absolute counts were still substantially greater than counts recovered from 
archaeological remains 
[@trompDietaryNondietary2015; @trompEDTACalculus2017; @wesolowskiEvaluatingMicrofossilContent2010], which could in part be due to the 
lack of oral amylase activity in our model. 
<!-- Scientific findings as extension of previous work -->
Previous research conducted on dental calculus from contemporary humans and non-human
primates suggest a high level of stochasticity involved in the retention of 
starch granules in dental calculus, and that starch granules extracted from dental 
calculus are underrepresented with regard
to actual starch intake, which is consistent with our findings (illustrated by high
standard deviations and low proportional incorporation).
Leonard and colleagues [-@leonardDentalCalculus2015] found individual
calculus samples to be a poor predictor of diet in a population, as many of the 
consumed plants were missing from some individual samples, but were present in others.   
Power and colleagues [-@powerChimpCalculus2015] 
presented similar findings in non-human primates, where phytoliths were more
representative of individual diets than starch granules.
The size bias is also consistent with the findings by Power and colleagues
[-@powerChimpCalculus2015],
who found that plants producing starches 10--20 $\mu$m in
size were over-represented; however, the representation of granules larger than
20 $\mu$m in their study is unclear.

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 $\mu$m in
maximum length were underrepresented in the calculus samples compared to the original 
starch solutions, an effect that was consistent across all three treatments. Medium
granules (10--20 $\mu$m) were often over-represented
(Table \@ref(tab:sample-prop-tbl), 
and 
Figure \@ref(fig:ratio-plots)).
Large potato granules were most affected, potentially because of the greater
size-range. They can reach up to 100 $\mu$m in maximum length, whereas wheat 
granules generally only reach up to 35 $\mu$m 
[@gismondiStarchGranulesData2019; @haslamDecompositionStarch2004; @seidemannStarchAtlas1966, 174-176].
Granule morphology may also play a role. Large wheat granules 
are lenticular and have a larger surface area compared 
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area 
[@vandeveldeStarchMorphology2002; @janeAnthologyStarch1994; @reichertStarchBible1913b, 364-365; @seidemannStarchAtlas1966, 174-176]. 
Another potentially important factor 
from our results is the size of the calculus deposit. We found a 
`r strength_cor` `r direct_cor` correlation between size of biofilm deposit and
retained starch granules (Figure \@ref(fig:cor-plot)A), meaning larger calculus
deposits contain a higher quantity of granules;
a result that contradicts findings from archaeological contexts
[@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeography2014].
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \@ref(fig:cor-plot)B).
While the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches. This may also explain
the lower mean retention of starch granules in mixed treatments compared to
wheat treatments. Wheat treatment samples (mean = `r subset(anova_tbl, treatment == "wheat")$mean` mg) were on average larger than mixed treatment samples (mean = `r subset(anova_tbl, treatment == "mix")$mean` mg) (Table \@ref(tab:anova-tbl)); and while mixed treatment solutions contained the highest mean overall
granule counts, wheat treatment samples had the highest mean starch retention.
Further research is needed to determine why this differs from previous archaeological
findings.

@@

<!-- limitations of the platform show which questions would be unanswered -->
The absence of $\alpha$-amylase in the model is a limitation of this study, as 
the total granule counts were not subject to hydrolysis. This would likely have 
reduced and affected the size ratios, as smaller starches may be more 
susceptible to hydrolysis 
[@francoStarchDegradation1992; @haslamDecompositionStarch2004]. The absence may
also affect biofilm growth due to the lack of amylase-bacterium interactions
[@nikitkovaStarchBiofilms2013]. Conversely, the model may benefit from the
absence of $\alpha$-amylase, because it can allow us to directly explore
its effect on starch counts in future experiments, where $\alpha$-amylase can be 
added to the model in concentrations similar to those found in the oral cavity
[@scannapiecoSalivaryAmylase1993]. 
We are able to show how absolute counts in the treatments cause a difference in
incorporation. However, this
was merely a side-effect of the difference in the number of granules in potato and 
wheat solutions of the same concentration (w/v). Further research should test 
multiple differing concentrations of the same starch type.
The use of EDTA may also have affected counts. While previous studies have shown 
negligible morphological changes caused by exposure to EDTA 
[@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
these studies have not considered changes to separate size
categories within starch types, and whether shifts in size ratios occur due to
exposure to the pre-treatment chemicals.
The total number of granules on a slide often exceeded a number that 
was feasible to count in a reasonable time period, so we calculated the total 
counts by extrapolating from three slide transects. 
Thus, we reasonably assume that the three transects are a good representation 
of the entire slide, and that the distribution of all granules on the slide is 
relatively homogeneous.   
Finally, we only used native starches in the experimental procedure and the results 
will likely differ for processed starches [@graneroStarchTaphonomy2020]. 
Based on the comparatively low counts obtained by Leonard
and colleagues [-@leonardDentalCalculus2015, Supplement 2], processing and amylase
may have a substantial effect on starch granule retention in the oral cavity.

>>>>

<!-- Big summary/important statement for the science -->
Here, we have provided a method for exploring the incorporation of dietary 
starches into the mineral matrix of a dental calculus biofilm model. Our results show
that a very low proportion of the starches exposed to the biofilm during growth are
retained in the mineral matrix, and that the size of the starch granules
may affect the likelihood of incorporation. The proportions of starch granules 
(of all sizes) present in the extracted samples were similar across all treatments
(`r paste(min(perc_tbl$total), max(perc_tbl$total), sep = " to ")`), 
despite large differences in absolute granule counts between wheat 
(mean = `r format(sol_comb_out$total[5], big.mark = ",")`) 
and potato
(mean = `r format(sol_comb_out$total[4], big.mark = ",")`)
solutions.  
The absolute counts, however, differed more visibly between treatments and was
proportional with the total count of granules in the treatment solutions. Wheat 
and mixed solutions had the highest absolute mean count of starch granules, and 
also had the highest absolute mean count of starch granules extracted from the 
dental calculus
(Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).
This suggests that the starches that are more frequently consumed will be present
in higher quantities in the dental calculus, at least prior to inhumation and 
degradation in the burial environment. 
Despite the low proportion of granules recovered from the model calculus 
(`r paste(min(perc_tbl$total), max(perc_tbl$total), sep = " to ")`),
the absolute counts were still substantially greater than counts recovered from 
archaeological remains 
[@trompDietaryNondietary2015; @trompEDTACalculus2017; @wesolowskiEvaluatingMicrofossil2010], which could in part be due to the 
lack of oral amylase activity in our model. 
<!-- Scientific findings as extension of previous work -->
Previous research conducted on dental calculus from contemporary humans and non-human
primates suggest a high level of stochasticity involved in the retention of 
starch granules in dental calculus, and that starch granules extracted from dental 
calculus are underrepresented with regard
to actual starch intake, which is consistent with our findings (illustrated by high
standard deviations and low proportional incorporation).
Leonard and colleagues [-@leonardPlantMicroremains2015] found individual
calculus samples to be a poor predictor of diet in a population, as many of the 
consumed plants were missing from some individual samples, but were present in others.   
Power and colleagues [-@powerChimpCalculus2015] 
presented similar findings in non-human primates, where phytoliths were more
representative of individual diets than starch granules.
The size bias is also consistent with the findings by Power and colleagues
[-@powerChimpCalculus2015],
who found that plants producing starches 10--20 $\mu$m in
size were over-represented; however, the representation of granules larger than
20 $\mu$m in their study is unclear.

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 $\mu$m in
maximum length were underrepresented in the calculus samples compared to the original 
starch solutions, an effect that was consistent across all three treatments. Medium
granules (10--20 $\mu$m) were often over-represented
(Table \@ref(tab:sample-prop-tbl), 
and 
Figure \@ref(fig:ratio-plots)).
Large potato granules were most affected, potentially because of the greater
size-range. They can reach up to 100 $\mu$m in maximum length, whereas wheat 
granules generally only reach up to 35 $\mu$m 
[@gismondiStarchGranules2019; @haslamDecompositionStarch2004; @seidemannStarchAtlas1966, 174-176].
Granule morphology may also play a role. Large wheat granules 
are lenticular and have a larger surface area compared 
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area 
[@vandeveldeStarchMorphology2002; @janeAnthologyStarch1994; @reichertStarchBible1913b, 364-365; @seidemannStarchAtlas1966, 174-176]. 
Another potentially important factor 
from our results is the size of the calculus deposit. We found a 
`r strength_cor` `r direct_cor` correlation between size of biofilm deposit and
retained starch granules (Figure \@ref(fig:cor-plot)A), meaning larger calculus
deposits contain a higher quantity of granules;
a result that contradicts findings from archaeological contexts
[@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeography2014].
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \@ref(fig:cor-plot)B).
While the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches. This may also explain
the lower mean retention of starch granules in mixed treatments compared to
wheat treatments. Wheat treatment samples (mean = `r subset(anova_tbl, treatment == "wheat")$mean` mg) were on average larger than mixed treatment samples (mean = `r subset(anova_tbl, treatment == "mix")$mean` mg) (Table \@ref(tab:anova-tbl)); and while mixed treatment solutions contained the highest mean overall
granule counts, wheat treatment samples had the highest mean starch retention.
Further research is needed to determine why this differs from previous archaeological
findings.

@@

<!-- limitations of the platform show which questions would be unanswered -->
The absence of $\alpha$-amylase in the model is a limitation of this study, as 
the total granule counts were not subject to hydrolysis. This would likely have 
reduced and affected the size ratios, as smaller starches may be more 
susceptible to hydrolysis 
[@francoStarchDegradation1992; @haslamDecompositionStarch2004]. The absence may
also affect biofilm growth due to the lack of amylase-bacterium interactions
[@nikitkovaStarchBiofilms2013]. Conversely, the model may benefit from the
absence of $\alpha$-amylase, because it can allow us to directly explore
its effect on starch counts in future experiments, where $\alpha$-amylase can be 
added to the model in concentrations similar to those found in the oral cavity
[@scannapiecoSalivaryAmylase1993]. 
We are able to show how absolute counts in the treatments cause a difference in
incorporation. However, this
was merely a side-effect of the difference in the number of granules in potato and 
wheat solutions of the same concentration (w/v). Further research should test 
multiple differing concentrations of the same starch type.
The use of EDTA may also have affected counts. While previous studies have shown 
negligible morphological changes caused by exposure to EDTA 
[@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
these studies have not considered changes to separate size
categories within starch types, and whether shifts in size ratios occur due to
exposure to the pre-treatment chemicals.
The total number of granules on a slide often exceeded a number that 
was feasible to count in a reasonable time period, so we calculated the total 
counts by extrapolating from three slide transects. 
Thus, we reasonably assume that the three transects are a good representation 
of the entire slide, and that the distribution of all granules on the slide is 
relatively homogeneous.   
Finally, we only used native starches in the experimental procedure and the results 
will likely differ for processed starches [@graneroStarchTaphonomy2020]. 
Based on the comparatively low counts obtained by Leonard
and colleagues [-@leonardPlantMicroremains2015, Supplement 2], processing and amylase
may have a substantial effect on starch granule retention in the oral cavity.

### 2023-05-31 analysis/paper/intro.Rmd
fix: broken citekeys

<!-- more detailed background, comprehensible to scientists in related disciplines -->
Until recently, only a few studies directly investigated the presence of plant
microremains
in the dental calculus of archaeological remains. The ability to extract phytoliths
from the dental calculus of archaeological fauna to investigate diet was first 
noted by Armitage [-@armitageExtractionIdentification1975], 
and later by Middleton and Rovner [-@middletonOpalPhytoliths1994],
and Fox and colleagues [-@foxPhytolithCalculus1996]. Starches and phytoliths were
extracted from human dental calculus by Cummings and Magennis [-@cummingsMayanCalculus1997].  
In more recent years, the study of dental calculus has increased exponentially, 
and the wealth of information contained within the mineralised matrix has largely
been acknowledged. The use of dental calculus spans a wide
variety of archaeological research areas, such as oral microbiome 
characterisation (including pathogens) through the analysis of DNA and proteins 
[@adlerSequencingAncient2013; @warinnerPathogensHost2014],
microbotanical remains [@henryCalculusSyria2008; @hardyStarchGranulesDental2009;
@mickleburghNewInsights2012], 
other organic residues and proteins from dietary compounds 
[@buckleyDentalCalculus2014; @hendyProteomicCalculus2018], 
and nicotine use [@eerkensDentalCalculus2018]. Especially the extraction 
of starch granules has become a rich source of dietary
information, as starch granules have proven to preserve well within dental calculus
over a variety of geographical and temporal ranges 
[@pipernoStarchGrains2008; @henryNeanderthalCalculus2014; @taoWheatCalculus2020;
@jovanovicNeolithicCalculus2021].

>>>>

<!-- more detailed background, comprehensible to scientists in related disciplines -->
Until recently, only a few studies directly investigated the presence of plant
microremains
in the dental calculus of archaeological remains. The ability to extract phytoliths
from the dental calculus of archaeological fauna to investigate diet was first 
noted by Armitage [-@armitageExtractionIdentification1975], 
and later by Middleton and Rovner [-@middletonOpalPhytoliths1994],
and Fox and colleagues [-@foxPhytolithCalculus1996]. Starches and phytoliths were
extracted from human dental calculus by Cummings and Magennis [-@cummingsMayanCalculus1997].  
In more recent years, the study of dental calculus has increased exponentially, 
and the wealth of information contained within the mineralised matrix has largely
been acknowledged. The use of dental calculus spans a wide
variety of archaeological research areas, such as oral microbiome 
characterisation (including pathogens) through the analysis of DNA and proteins 
[@adlerSequencingAncient2013; @warinnerPathogensHost2014],
microbotanical remains [@henryCalculusSyria2008; @hardyStarchGranules2009;
@mickleburghNewInsights2012], 
other organic residues and proteins from dietary compounds 
[@buckleyDentalCalculus2014; @hendyProteomicCalculus2018], 
and nicotine use [@eerkensDentalCalculus2018]. Especially the extraction 
of starch granules has become a rich source of dietary
information, as starch granules have proven to preserve well within dental calculus
over a variety of geographical and temporal ranges 
[@pipernoStarchGrains2008; @henryNeanderthalCalculus2014; @taoWheatCalculus2020;
@jovanovicNeolithicCalculus2021].

### 2023-05-31 analysis/paper/matmet.Rmd
fix: broken citekeys

In this study we employ a multispecies oral biofilm model following a modified
protocol from Sissons and colleagues [-@sissonsMultistationPlaque1991] and
Shellis [-@shellisSyntheticSalivaCultural1978]. In brief, a biofilm inoculated
with whole saliva was grown on a substrate suspended in artificial saliva, and
fed with sugar (sucrose). After several days of growth, the biofilm was exposed
to starch solutions. Mineralisation of the biofilm was aided by exposure to a
calcium phosphate solution. After 25 days of growth, the mineralised biofilm was
collected for further analysis. The setup comprises a polypropylene 
24 deepwell PCR plate (KingFisher 97003510) with a lid containing 24 pegs, which 
is autoclaved at 120&deg;C, 1 bar overpressure, for 20 mins. The individual pegs 
were the substrata on which the calculus grew. Using this system allowed for easy
transfer of the growing biofilm between saliva, feeding solutions, 
and mineral solutions.

The artificial saliva (AS) is a modified version of the basal medium mucin (BMM)
described by Sissons and colleagues [-@sissonsMultistationPlaque1991]. 
It contains 2.5 g/l partially purified mucin from porcine stomach (Type III, Sigma M1778), 
5 g/l trypticase peptone (Roth 2363.1), 10 g/l proteose peptone (Oxoid LP0085),
5 g/l yeast extract (BD 211921), 2.5 g/l KCl, 0.35 g/l NaCl, 1.8 mmol/l CaCl~2~, 
5.2 mmol/l Na~2~HPO~4~ [@sissonsMultistationPlaque1991], 6.4 mmol/l NaHCO~3~ 
[@shellisSyntheticSalivaCultural1978], 2.5 mg/l haemin. This is subsequently 
adjusted to pH 7 with NaOH pellets and stirring, autoclaved (15 min, 120&deg;C, 
1 bar overpressure), and supplemented with 5.8 $\mu$mol/l menadione, 5 mmol/l urea, 
and 1 mmol/l arginine [@sissonsMultistationPlaque1991].

@@

After 15 days, mineralisation was encouraged with a 
calcium phosphate monofluorophosphate urea (CPMU) solution containing 
20 mmol/l CaCl~2~, 12 mmol/l NaH~2~PO~4~, 5 mmol/l Na~2~PO~3~F, 500 mmol/l urea, 
and (0.04 g/l MgCl)
[@pearceConcomitantDepositionStrontium1987; @sissonsMultistationPlaque1991].
The substrata were submerged in 1 ml/well CPMU for six minutes, five times 
daily, in a two-hour cycle. During the mineralisation period, starch treatments
were reduced to once per day after the five CPMU treatments. This process was repeated 
for 10 days until the end of the experiment on day 24 
(see Figure \@ref(fig:protocol-fig) for an overview of the protocol).

>>>>

In this study we employ a multispecies oral biofilm model following a modified
protocol from Sissons and colleagues [-@sissonsMultistationPlaque1991] and
Shellis [-@shellisSyntheticSaliva1978]. In brief, a biofilm inoculated
with whole saliva was grown on a substrate suspended in artificial saliva, and
fed with sugar (sucrose). After several days of growth, the biofilm was exposed
to starch solutions. Mineralisation of the biofilm was aided by exposure to a
calcium phosphate solution. After 25 days of growth, the mineralised biofilm was
collected for further analysis. The setup comprises a polypropylene 
24 deepwell PCR plate (KingFisher 97003510) with a lid containing 24 pegs, which 
is autoclaved at 120&deg;C, 1 bar overpressure, for 20 mins. The individual pegs 
were the substrata on which the calculus grew. Using this system allowed for easy
transfer of the growing biofilm between saliva, feeding solutions, 
and mineral solutions.

The artificial saliva (AS) is a modified version of the basal medium mucin (BMM)
described by Sissons and colleagues [-@sissonsMultistationPlaque1991]. 
It contains 2.5 g/l partially purified mucin from porcine stomach (Type III, Sigma M1778), 
5 g/l trypticase peptone (Roth 2363.1), 10 g/l proteose peptone (Oxoid LP0085),
5 g/l yeast extract (BD 211921), 2.5 g/l KCl, 0.35 g/l NaCl, 1.8 mmol/l CaCl~2~, 
5.2 mmol/l Na~2~HPO~4~ [@sissonsMultistationPlaque1991], 6.4 mmol/l NaHCO~3~ 
[@shellisSyntheticSaliva1978], 2.5 mg/l haemin. This is subsequently 
adjusted to pH 7 with NaOH pellets and stirring, autoclaved (15 min, 120&deg;C, 
1 bar overpressure), and supplemented with 5.8 $\mu$mol/l menadione, 5 mmol/l urea, 
and 1 mmol/l arginine [@sissonsMultistationPlaque1991].

@@

After 15 days, mineralisation was encouraged with a 
calcium phosphate monofluorophosphate urea (CPMU) solution containing 
20 mmol/l CaCl~2~, 12 mmol/l NaH~2~PO~4~, 5 mmol/l Na~2~PO~3~F, 500 mmol/l urea, 
and (0.04 g/l MgCl)
[@pearceConcomitantDeposition1987; @sissonsMultistationPlaque1991].
The substrata were submerged in 1 ml/well CPMU for six minutes, five times 
daily, in a two-hour cycle. During the mineralisation period, starch treatments
were reduced to once per day after the five CPMU treatments. This process was repeated 
for 10 days until the end of the experiment on day 24 
(see Figure \@ref(fig:protocol-fig) for an overview of the protocol).
